Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-01-28 CAV 2025 TODO/Paper_noProof.tex
work
comparison
formula
exp

\newcommand{\val}{{\textrm{value}}}
\newcommand{\Val}{{\textrm{value}}}
\newcommand{\MILP}{{\textrm{MILP}}}
\newcommand{\LP}{{\textrm{LP}}}
\newcommand{\Improve}{\mathrm{Improve}}
\newcommand{\Utility}{\mathrm{Utility}}
\newcommand{\Sol}{\mathrm{Sol}}
\newcommand{\sol}{\mathrm{sol}}

@@

\section{Conclusion}
{\color{blue}
	In this paper, we developed a novel Utility function to select few ReLU nodes to consider with binary variables to compute accurately bounds on neurons of DNNs. The novelties are that it focuses on {\em improvement} wrt a given input, rather than on generic {\em sensitivity} of a neuron wrt to a ReLU node, and it uses the solution of one call to an (efficient LP) solver to evaluate this improvement. This makes the choice particularly efficient, necessitating $\approx4$x less integer variables than previous proposals \cite{DivideAndSlide} for the same accuracy.} 
Our empirical studies revealed that this can yield highly accurate results, verifying up to 40\% more images than the SOTA ($\alpha,\beta$-Crown, winner of the 4 last VNNComp), with the same runtime, for DNNs with up to 20 000 neurons. 
The reason is that $\alpha,\beta$-Crown hits a complexity barrier, similarly as other competing solutions, when considering hard (even small) DNNs. This opens a lot of perspectives, among which: verifying efficiently other hard instances; certifying $\epsilon$-robustness of images for $\epsilon$ as large as possible; 
verifying global rather than local properties \cite{lipshitz}.

>>>>

\newcommand{\val}{{\textrm{value}}}
\newcommand{\Val}{{\textrm{value}}}
\newcommand{\MILP}{{\textrm{MILP}}}
\newcommand{\LP}{{\textrm{LP}}}
\newcommand{\Improve}{\mathrm{Improve}}
\newcommand{\Utility}{\mathrm{SAS}}
\newcommand{\Sol}{\mathrm{Sol}}
\newcommand{\sol}{\mathrm{sol}}

@@

\section{Conclusion}
In this paper, we developed a novel solution-aware scoring (SAS) function to select few ReLU nodes to consider with binary variables to compute accurately bounds in DNNs. 
The solution awareness allows SAS to compute an accurate score for each ReLU, which enables partial MILP to be very efficient, necessitating $\approx4$x less binary variables than previous proposals \cite{DivideAndSlide} for the same accuracy. As the worst-case complexity is exponential in the number of binary variables, this big leap ahead has large implication in terms of scalability to larger DNNs, making it possible to verify quite large DNNs such as CNN-B-Adv with 2M parameters and 20000 activation functions.

While $\alpha,\beta$-Crown is known to be extremely efficient to solve easier verification instances, we exhibit many cases (complex instances) where its worst-case exponential complexity in the number of ReLUs is tangible, with unfavorable scaling (Table \ref{table_beta}). Resorting to Hybrid MILP, a divide-and-conquer approach \cite{DivideAndSlide}, revisited thanks to the very efficient solution-aware scoring, revealed to be a much better trade-off than augmenting $\alpha,\beta$-Crown time-outs, with $8\%$ to $40\%$ less undecided images at ISO runtime.

Last, we adapted BaB-SR \cite{BaB} and FSB \cite{FSB} as global scoring (GS) functions. 
We compare them both theoretically and experimentally with solution-aware scoring: While GS are not as accurate as SAS for ReLU {\em selection}, they are better suited to select the {\em order} with which to branch over this selection, at least when the order is static and does not depend upon the branch. This opens up interesting future research directions, to hybridize different scoring for different tasks;  e.g. verifying global \cite{lipshitz}, \cite{sensing} rather than local (robustness) properties.

### 2025-01-28 CAV 2025 TODO/experiments.tex
comp
exp

%
We implemented \toolname\ in Python 3.8, and Gurobi 9.52 was used for solving LP and MILP problems. We conducted our evaluation on an AMD Threadripper 7970X  ($32$ cores$@4.0$GHz) with 256 GB of main memory and 2 NVIDIA RTX 4090. The  objectives of our evaluation was to answer the following  questions:

@@

We also experimented in Table \ref{table.order} on fixing an ordering on variable in Gurobi, with a fixed selection of nodes by SAS. Indeed, Gurobi relies on a BaB procedure to compute bounds. The standard Gurobi procedure is adaptative in each branch. We compared vs the fix order provided by SAS, as well as the fixed order provided by FSB, for both accuracy and runtime (lower is better).

\begin{table}[h!]	
	\centering
	\begin{tabular}{|c||c c|c c|c c|}
		\hline
		  & \multicolumn{2}{c|}{30 open ReLUs} & \multicolumn{2}{c|}{45 open ReLUs} & \multicolumn{2}{c|}{60 open ReLUs} \\ 
		 order & uncert & time & uncert & time & uncert & time \\
		\hline \hline
		Gurobi &   3.987  & 260s  & 4.016 & 266s& 4.053 & 262s\\ 
		SAS &   3.989  & 264s  & 4.031 & 266s& 4.060 & 263s\\
		FSB &   3.988  & {\bf 246s}  & 4.018 & {\bf 248s}& 4.033 & {\bf 248s}\\ \hline
	\end{tabular}
	\caption{Comparison of different ordering with selection of ReLUs by SAS.}
	\label{table.order}
	%\vspace{-0.6cm}
\end{table}

Here,  SAS is as efficient as Gurobi, although the ordering is fixed, meaning that the ordering compensates for the non adaptivity. However, the fixed GS ordering (with the SAS selection) is even better, with slightly better accuracy and lower runtime. As explained earlier, this is due to the fact that global ordering is preferable when branching on choices far from the optimal solution.

@@

We conducted our evaluation on the neural networks tested in \cite{crown}
which display a large number ($\geq 20\%$) of images undecided  by $\alpha,\beta$-Crown. That is, these DNNs are {\em hard} to verify. Namely, these are 6 ReLU-DNNs: 5 MNIST DNN that can be found in the \href{https://github.com/eth-sri/eran}{ERAN GitHub} 
(the 4th to the 8th DNNs provided) as well as 1 CIFAR CNN from 
\cite{AdversarialTrainingAndProvableDefenses}, see also \cite{SDPFI}, 
which can be downloaded from the \href{https://github.com/Verified-Intelligence/alpha-beta-CROWN}{$\alpha,\beta$-Crown GitHub}. 
We commit to the same $\epsilon$ settings as in \cite{crown}, that are recalled in Table \ref{table_beta}. 
For reference, we also report an easy but very large ResNet Network for CIFAR10, already tested with $\alpha,\beta$ CROWN. 
We report in Table \ref{table_hybrid} the $\%$ of undecided images, that is the $\%$ of images than can be neither falsified (by $\alpha,\beta$-CROWN) nor verified by the tested verifier, among the 100 first images for each MNIST or CIFAR10 benchmark. 
The exact same DNNs and $\epsilon$ are used in Tables \ref{table_beta} and \ref{table_hybrid}.

\begin{table}[t!]
	\centering
	\begin{tabular}{||l||c|c|c||c||c||}
		\hline \hline
		 & $\alpha,\beta$-Crown & $\alpha,\beta$-Crown & $\alpha,\beta$-Crown & Refined & \bf Hybrid \\ 
		 Network & TO=10s & TO=30s & TO=2000s & $\beta$-Crown & \bf MILP \\ 
		\hline
		MNIST 5$\times$100 & 57\% (6.9s) & 55\% (18.9s) & 50\% (1026s) & 13\% 
		(92s) & 13\% \bf (46s) \\ \hline
		MNIST 5$\times$200 & 50\% (6.5s) & 47\% (17s) & 46\% (930s) & 9\% (80s) & \bf 8\% (71s) \\ \hline
		MNIST 8$\times$100 & 63\% (7.2s) & 58\% (20s) & 58\% (1163s) & 21\% (102s) & \bf 15\% (61s) \\ \hline
		MNIST 8$\times$200 & 56\% (6.8s) & 55\% (18s) & 54\% (1083s) & 16\% (83s) & \bf 8\% (78s) \\ \hline
		MNIST 6$\times$500 & 53\% (6.4s) & 51\% (16s) & 50\% (1002s) & $-$ & 
		\bf 10\% (402s) \\ \hline
		CIFAR CNN-B-adv & 28\% (4.3s) & 22\% (8.7s) & 20\% (373s) & $-$ & \bf 11\% (417s) \\ \hline \hline
		CIFAR ResNet & 0\% (2s) & 0\% (2s) & 0\% (2s) & $-$ & 0\% (2s) \\ \hline \hline
	\end{tabular}
	\caption{Undecided images ($\%$, {\em lower is better}) as computed by $\alpha,\beta$-Crown, Refined $\beta$-Crown and Hybrid MILP on 7 DNNs (average runtime per image). The 6 first DNNs are hard instances. The last DNN (ResNet) is an easy instance (trained using Wong to be easy to verify, but with a very low accuracy level), provided for reference.}
	\label{table_hybrid}
	\end{table}

{\em Analysis}: on easy instances, Hybrid MILP is virtually similar to $\alpha,\beta$-CROWN, 
as it is called first and it is sufficient to have $0\%$ undecided images, as shown even on the very large ResNet.

On hard instances (the 6 first DNNs tested), Hybrid MILP is very accurate, only leaving $8\%$-$15\%$ of images undecided, with runtime taking less than $500s$ in average per image, and even 10 times less on smaller DNNs. It can scale up to quite large hard DNNs, such as CNN-B-Adv with 2M parameters and 20K activations.

Compared with $\alpha,\beta$-Crown with a time-out of TO=2000s, it is much more accurate, with a reduction of undecided images by $9\%-43\%$. 
It is also from 20x faster on smaller networks to similar time on the largest DNN.
Compared with $\alpha,\beta$-Crown with a time-out of TO=30s, the accuracy gap is even larger (e.g. $11\%$ for CNN-B-Adv, i.e. half undecided images), although the average runtime is also obviously larger (solving hard istances takes longer than solving easy instances).

Last, compared with {\em Refined} $\beta$-Crown, we can observe three patterns: on the shallowest DNNs (5$\times$100, 5$\times$200), Refined $\beta$-Crown can run full MILP on almost all nodes, reaching almost the same accuracy than  Hybrid MILP, but with longer runtime (up to 2 times on 5$\times$100). On intermediate DNNs (8$\times$100, 8$\times$200), full MILP invoked by Refined $\beta$-Crown can only be run on a fraction of the neurons, and the accuracy is not as good as Hybrid MILP, with $6\%-8\%$ more undecided images (that is double on 8$\times$200), while having longer runtime. Last but not least, Refined $\beta$-Crown cannot scale to larger instances (6$\times$500, CNN-B-Adv), while Hybrid MILP can.

@@

We voluntarily limited the comparison to $\alpha,\beta$-Crown because it is the most efficient verifier to date, and to consider a spectre of parameters to understand $\alpha,\beta$-Crown scaling without too much clutter.

Notice that results for other verifiers (PRIMA \cite{prima}, SDP-FO \cite{SDPFI}, etc) were already reported on these DNNs \cite{crown}, with unfavorable comparison vs $\alpha,\beta$-CROWN. Further, we reported accuracy of complete verifiers, NNenum \cite{nnenum}, Marabou \cite{Marabou,Marabou2}, respectively 4th, 3rd 
of the last VNNcomp'24 \cite{VNNcomp24}, as well as full MILP \cite{MILP} in Table \ref{table_complete}, showing that these verifiers are not competitive on hard instances either, even on the smallest hard DNN. Concerning MnBAB \cite{ferrari2022complete}, we tested it in appendix, and it compares slightly unfavorably in time and accuracy towards $\alpha,\beta$-CROWN on CNN-B-Adv and {\em hard} MNIST DNNs at several time-out settings. 
Last, Pyrat \cite{pyrat} (2sd in the latest VNNComp) is not open source, which made running it impossible.

>>>>

%
We implemented \toolname\ in Python 3.8: it first calls $\alpha,\beta$-Crown with small time out (10s for small and 30s for larger DNNs), and then call pMILP on the undecided inputs.
Gurobi 9.52 was used for solving LP and MILP problems. We conducted our evaluation on an AMD Threadripper 7970X  ($32$ cores$@4.0$GHz) with 256 GB of main memory and 2 NVIDIA RTX 4090. The  objectives of our evaluation was to answer the following  questions:

@@

Gurobi relies on a BaB procedure to compute bounds. The standard Gurobi ordering of variables is adaptative in each branch. We report in Table \ref{table.order} experiments on changing the ordering on variables in Gurobi, when the selection of ReLUs is fixed by SAS. We compared it with the static order provided by SAS, as well as the static order provided by GS, for runtime at a fixed accuracy (MIPGAP=0.01, so $1\%$ of difference in uncertainty is possible).
The test are done on the same CIFAR image 85  and CNN-B-Adv.

\begin{table}[t]	
	\centering
	\begin{tabular}{|c||c c|c c|c c|c c|}
		\hline
		  & \multicolumn{2}{c|}{45 open ReLUs} & \multicolumn{2}{c|}{50 open ReLUs} 
		  & \multicolumn{2}{c|}{55 open ReLUs} & \multicolumn{2}{c|}{60 open ReLUs}\\ 
		 order & uncert & time & uncert & time & uncert & time & uncert & time \\
		\hline \hline
		Gurobi &  & 252s & &  & & &   & 306s  \\ 
		SAS (static) &  & 448s & &  & & &   & 971s \\
		GS (static) & & {\bf 251s} & &  &  & &  & {\bf 281s} \\ \hline
	\end{tabular}
	\caption{Comparison of different orderings with selection of ReLUs by SAS.}
	\label{table.order}
	%\vspace{-0.6cm}
\end{table}

SAS ordering is particularly counter-productive because it is static, and it is accurate only for one branch (the most complicated one), whereas Gurobi can adapt the order to each branch.
However, the GS ordering (with the SAS selection) is better as it is general and not local to a solution, with better runtime than Gurobi, although it is static whereas Gurobi order is adaptative. Using SAS order for the solution branch, and GS ordering for other branches should be more efficient, but we cannot experiment that within the closed source Gurobi solver.

@@

To assess the verification efficiency and runtime vs $\alpha,\beta$-Crown, we 
conducted evaluations on the neural networks which are {\em complex} to verify, from \cite{crown}. Namely, these are 6 ReLU-DNNs: 5 MNIST DNN that can be found in the \href{https://github.com/eth-sri/eran}{ERAN GitHub} 
(the 4th to the 8th DNNs provided) as well as 1 CIFAR CNN from 
\cite{AdversarialTrainingAndProvableDefenses}, see also \cite{SDPFI}, 
which can be downloaded from the \href{https://github.com/Verified-Intelligence/alpha-beta-CROWN}{$\alpha,\beta$-Crown GitHub}. 
We commit to the same $\epsilon$ settings as in \cite{crown}, that are recalled in Table \ref{table_beta}. 
For reference, we also report an easy but very large ResNet Network for CIFAR10, already tested with $\alpha,\beta$ CROWN. 
We report in Table \ref{table_hybrid} the $\%$ of undecided images, that is the $\%$ of images than can be neither falsified (by $\alpha,\beta$-CROWN) nor verified by the tested verifier, among the 100 first images for each MNIST or CIFAR10 benchmark. 
The exact same DNNs and $\epsilon$ are used in Tables \ref{table_beta} and \ref{table_hybrid}.

\begin{table}[t!]
	\centering
	\begin{tabular}{||l||c|c|c||c||c||}
		\hline \hline
		 & $\alpha,\beta$-Crown & $\alpha,\beta$-Crown & $\alpha,\beta$-Crown & Refined & \bf Hybrid \\ 
		 Network & TO=10s & TO=30s & TO=2000s & $\beta$-Crown & \bf MILP \\ 
		\hline
		MNIST 5$\times$100 & 57\% (6.9s) & 55\% (18.9s) & 50\% (1026s) & 13\% 
		(92s) & 13\% \bf (46s) \\ \hline
		MNIST 5$\times$200 & 50\% (6.5s) & 47\% (17s) & 46\% (930s) & 9\% (80s) & \bf 8\% (71s) \\ \hline
		MNIST 8$\times$100 & 63\% (7.2s) & 58\% (20s) & 58\% (1163s) & 21\% (102s) & \bf 15\% (61s) \\ \hline
		MNIST 8$\times$200 & 56\% (6.8s) & 55\% (18s) & 54\% (1083s) & 16\% (83s) & \bf 8\% (78s) \\ \hline
		MNIST 6$\times$500 & 53\% (6.4s) & 51\% (16s) & 50\% (1002s) & $-$ & 
		\bf 10\% (402s) \\ \hline
		CIFAR CNN-B-adv & 28\% (4.3s) & 22\% (8.7s) & 20\% (373s) & $-$ & \bf 11\% (417s) \\ \hline \hline
		CIFAR ResNet & 0\% (2s) & 0\% (2s) & 0\% (2s) & $-$ & 0\% (2s) \\ \hline \hline
	\end{tabular}
	\caption{Undecided images ($\%$, {\em lower is better}) as computed by $\alpha,\beta$-Crown, Refined $\beta$-Crown and Hybrid MILP on 7 DNNs (average runtime per image). The 6 first DNNs are hard instances. The last DNN (ResNet) is an easy instance (trained using Wong to be easy to verify), provided for reference.}
	\label{table_hybrid}
	\end{table}

{\em Analysis}: overall, Hybrid MILP is very accurate, only leaving $8\%$-$15\%$ of images undecided, with runtime taking less than $500s$ in average per image, and even 10 times less on smaller DNNs. It can scale up to quite large hard DNNs, such as CNN-B-Adv with 2M parameters and 20K activations.

Compared with $\alpha,\beta$-Crown:
on easy instances (see the very large ResNet), Hybrid MILP is virtually similar to $\alpha,\beta$-CROWN, as $\alpha,\beta$-CROWN is called first because it is very efficient for such instances.

On hard instances (the 6 first DNNs tested), compared with $\alpha,\beta$-Crown with a time-out of TO=2000s, it is much more accurate, with a reduction of undecided images by $9\%-43\%$. It is also from 20x faster on smaller networks to similar time on the largest DNN.
Compared with $\alpha,\beta$-Crown with a time-out of TO=30s, the accuracy gap is even larger (e.g. $11\%$ for CNN-B-Adv, i.e. half undecided images), although the average runtime is also obviously larger (solving hard istances takes longer than solving easy instances).

Last, compared with {\em Refined} $\beta$-Crown, we can observe three patterns: on the shallowest DNNs (5$\times$100, 5$\times$200), Refined $\beta$-Crown can run full MILP on almost all nodes, reaching almost the same accuracy than  Hybrid MILP, but with longer runtime (up to 2 times on 5$\times$100). As size of DNNs grows (8$\times$100, 8$\times$200), full MILP invoked by Refined $\beta$-Crown can only be run on a fraction of the neurons, and the accuracy is not as good as Hybrid MILP, with $6\%-8\%$ more undecided images (that is double on 8$\times$200), while having longer runtime. Last but not least, Refined $\beta$-Crown cannot scale to larger instances (6$\times$500, CNN-B-Adv), while Hybrid MILP can.

@@

We voluntarily limited the comparison to $\alpha,\beta$-Crown because it is one of the most efficient verifier to date, which allowed us to consider a spectrum of parameters to understand $\alpha,\beta$-Crown scaling without too much clutter.

Interestingly, GCP-CROWN \cite{cutting} is slightly more accurate than $\alpha,\beta$-Crown on the DNNs we tested, but necessitates IBM CPLEX solver, which is not available to us.
We report result from \cite{cutting} pertinent to the DNNs we tested, and copy the comparison with $\alpha,\beta$-Crown. While results are similar to ours, they do not use exactly the same number of input images than in our test, and thus it is not directly comparable. 
On CNN-B-Adv, GCP-CROWN is $2\%$ more accurate (with higher runtime) than 
$\alpha,\beta$-Crown with 90s TO, so it is safe to consider that Hybrid MILP, which is $9\%$ more accurate than $\alpha,\beta$-Crown with 2000s TO,  is significantly more accurate than GCP-CROWN on this quite large DNN.

Notice that results for other verifiers (PRIMA \cite{prima}, SDP-FO \cite{SDPFI}, etc) were already reported on these DNNs \cite{crown}, with unfavorable comparison vs $\alpha,\beta$-CROWN. Further, we reported accuracy of NNenum \cite{nnenum}, Marabou \cite{Marabou,Marabou2}, respectively 4th, 3rd of the last VNNcomp'24 \cite{VNNcomp24}, as well as full MILP \cite{MILP} in Table \ref{table_complete}, showing that these verifiers are not competitive on hard instances either, even on the smallest hard DNN. Concerning MnBAB \cite{ferrari2022complete}, we tested it in appendix, and it compares slightly unfavorably in time and accuracy towards $\alpha,\beta$-CROWN on CNN-B-Adv and {\em hard} MNIST DNNs at several time-out settings. Last, Pyrat \cite{pyrat} (2sd in the latest VNNComp) is not open source, which made running it impossible. Details can be found in the appendix. Also, ablation studies showing what each feature of Hybrid MILP is bringing can be found in the appendix.
